---
title: "p8105_hw2_tp2806"
author: "Tejashree Prakash"
date: 2025-09-29
output: github_document
---

```{r setup, include=FALSE}
#Load libraries and data
library(tidyverse)
library(readxl)
library(lubridate)
```

## Problem 1 

Downloading and cleaning data. 
```{r}
#Download and clean pols data
pols =
  read_csv("data/fivethirtyeight_datasets/pols-month.csv") %>%
  separate(mon, into = c("year", "month", "day"), sep = "-", convert = TRUE) %>%
  mutate(month = month.name[month]) %>% #Convert numbers to month names 
  mutate(president = case_when( #Create case to show gop or dem president 
    prez_gop == 1 ~ "gop",
    prez_dem == 1 ~ "dem"
  )) %>%
  select(-prez_dem, -prez_gop, -day)

#Download and clean snp data 
snp = 
  read_csv("data/fivethirtyeight_datasets/snp.csv") %>%
  mutate(
    date = mdy(date) #turn into as.date format 
  ) %>%
  separate(date, into = c("year", "month", "day"), sep = "-", convert = TRUE) %>%
  mutate(month = month.name[month]) %>%
  select(year, month, close, -day)

#Download and clean unemployment data
unemployment = 
  read_csv("data/fivethirtyeight_datasets/unemployment.csv") %>%
  pivot_longer(
    cols = Jan:Dec,
    names_to = "month",
    values_to = "unemployment_percentage"
  ) %>% 
  mutate(
    month = month.name[match(month, month.abb)] #change abbreviated month names to full month names
  ) %>%
  rename(year = Year) #make the capitalization of columns consistent 
```

Merging all the datasets together. 

```{r}
#Merge datasets - starting with snp into pols 
merge <- 
  left_join(pols, snp, by = join_by(year, month)) #want to join by both year and month

final_merge <- 
  left_join(merge, unemployment, by = join_by(year, month)) 

knitr::kable(head(final_merge,15))
```

The 'snp' dataset contains 787 observations and 2 variables - the date and the closing values of the S&P stock index on that date. The 'unemployment' dataset was converted to long form, now containing unemployment percentage according to each year and month. The 'pols' dataset contains 822 observations and 9 variables - the key variables include month, year, and party of the president elect. The resulting dataset, after merging, reflects the base structure for the pols dataset, with the unique columns from 'snp' and 'unemployment' added onto it based on matching month and year. Therefore, any values from 'snp' and 'unemployment' that did not have a corresponding month/year with 'pols' was not included in the final dataset. The years in this final dataset range from `r min(final_merge$year)` to `r max(final_merge$year)`.  


## Problem 2

Downloading and cleaning data. 
```{r}
#Download and clean Mr. Trash Wheel dataset
mr_trash <- 
  read_excel("data/trash_wheel_data_update.xlsx", 
             sheet = "Mr. Trash Wheel",
             range = "A2:N710") %>%
  janitor::clean_names() %>%
  filter(!is.na(dumpster)) %>% #remove the rows that don't have dumpster info
  mutate(
    sports_balls = as.integer(round(sports_balls)) #convert sports balls column 
  ) %>%
  mutate(
    identity = "mr", #tag what original dataframe is
    year = as.numeric(year)
  )
knitr::kable(head(mr_trash,10))

#Download and clean Professor Trash Wheel dataset
prof_trash <- 
  read_excel("data/trash_wheel_data_update.xlsx", 
             sheet = "Professor Trash Wheel",
             range = "A2:M123") %>%
  janitor::clean_names() %>%
  filter(!is.na(dumpster)) %>%
  mutate(
    identity = "prof", #tag what original dataframe is
    year = as.numeric(year)
  )
knitr::kable(head(prof_trash,10))

#Download and clean Gwynns Falls Trash Wheel dataset
gwynn_trash <- 
    read_excel("data/trash_wheel_data_update.xlsx", 
             sheet = "Gwynns Falls Trash Wheel",
             range = "A2:L352") %>%
  janitor::clean_names() %>%
  filter(!is.na(dumpster)) %>%
  mutate(
    identity = "gwynn", #tag what original dataframe is
    year = as.numeric(year)
  )
knitr::kable(head(gwynn_trash,10))
```


Merging the trash datasets. 
```{r}
#Combine gwyn and prof trash first as they share the same columns. 
trash_wheels <- bind_rows(mr_trash, prof_trash, gwynn_trash)

#Calculate total weight of trash by prof trash
totalweightprof <- prof_trash %>%
  summarise(total_weight = sum(weight_tons, na.rm = TRUE))

#Calculate total cigarette butts by Gwynnda in June 2022
totalcig_gwyn <- gwynn_trash %>%
  filter(year == 2022, month == "June") %>%
  summarise(total_cigs = sum(cigarette_butts, na.rm = TRUE))

knitr::kable(head(trash_wheels,15))
```

After cleaning the data sets, I noticed that not all of the datasets had the same columns. For example, "Mr. Trash Wheel" contains the column "sports balls", whereas the other two datasets do not contain that variable. The merged dataset has `r nrow(trash_wheels)` observations. Key variables include: weight of trash collected, cigarette butts collected, plastic bottles, and the original dataset identifier. The total weight of trash collected by Professor Trash Wheel is `r totalweightprof$total_weight`. The total cigarette butts collected by Gwynnda in June 2022 is `r totalcig_gwyn$total_cigs`.


## Problem 3

```{r}

```


